---
title: "Reproducible Research: Course Project 1"
output: 
  html_document: 
    keep_md: yes
---

Project by Charles-Philippe Lajoie, as part of the Coursera Data Scientist Reproducible Research course.

The project looks at a CSV file containing the number of steps taken by a person (monitored with some personal device) as a function of time (daily 5-minute intervals) for two months. 

## Loading & preprocessing the data
The data contains 3 colums: steps, date, and interval. The CSV file is read in R and any rows with NA are, for the moment, discarded. Strings representing dates are also converted to R dates variables.
```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(chron)

rawtable=read.csv("activity.csv")

mytable <- filter(rawtable, is.na(steps)==FALSE)
mytable$date <- as.Date(mytable$date, format="%Y-%m-%d" )
```
 
## What is mean total number of steps taken per day?
```{r, echo=TRUE}
#GROUP DATA BY DATE:
daily <- group_by(mytable, date)
totalSteps <- summarise(daily, sumSteps=sum(steps))

meanSteps=mean(  totalSteps$sumSteps)
mediSteps=median(totalSteps$sumSteps)
print(meanSteps)
```
```{r, echo=FALSE}
nbins=max(totalSteps$sumSteps)/1000

hist( totalSteps$sumSteps, breaks=nbins, 
      col="red", xlab='Daily # of steps', 
      ylim=c(0,12), main="Total Daily Number of Steps")
legend(500, 10, legend = c(paste("Mean =", round(meanSteps, 1)),
                            paste("Median =",round(mediSteps, 1))), bty = "n")
```


## What is the average daily activity pattern?

```{r echo=FALSE}
date_format_tz <- function(format = "%Y-%m-%d", tz = "UTC") {
  function(x) format(x, format, tz=tz)
}
```

```{r}
mytable$interval <- strptime(sprintf("%04d", as.numeric(mytable$interval)), format="%H%M")
mytable$interval <- as.POSIXct(mytable$interval, format="%H:%M:%S")

inter <- group_by(mytable, interval)
meanSteps <- summarise(inter, meanSteps=mean(steps))

timemax <- meanSteps$interval[which.max(meanSteps$meanSteps)]

g <- ggplot(meanSteps, aes(interval, meanSteps)) + geom_point() + geom_line()
g <- g + labs(title = "Average # steps as a function of time in day") 
g <- g + labs(x = "Time", y = "Average # of steps")
g <- g + scale_x_datetime(date_breaks="2 hour", labels=date_format_tz("%H:%M", tz="EST"))
print(g)
```

The time at which the maximum average number of steps occured on `r timemax`.


## Imputing missing values
```{r}
missing <- filter(rawtable, is.na(steps)==TRUE)
nrow_missing <- (nrow(missing))
```
There are `r nrow_missing` rows with values NA in the original dataset.

Here, we are filling the NA values for a given day interval with the median of that interval for all 60 days (or so).
```{r echo=FALSE}
# IGNORE NAs FOR NOW & GROUP BY INTERVAL
# AND CALCULATE DAILY AVERAGE:
cleantable <- filter(rawtable, is.na(steps)==FALSE)
inter <- group_by(cleantable, interval)
meanSteps <- summarise(inter, meanSteps=median(steps))
```
```{r}
for(i in 1:nrow(missing)){
  int <- missing$interval[i]
  missing$steps[i] <- meanSteps$meanSteps[meanSteps$interval==int]
}

allvalues <- rbind(cleantable, missing)
allvalues$date <- as.Date(allvalues$date, format="%Y-%m-%d" )

# GROUP BY DATE & CALCULATE DAILY AVERAGE:
date_groups <- group_by(allvalues, date)
all <- summarise(date_groups, sumSteps=sum(steps))

meanSteps=mean(  all$sumSteps)
mediSteps=median(all$sumSteps)

print(meanSteps)
```

```{r echo=FALSE}
nbins=max(all$sumSteps)/1000
hist( all$sumSteps, breaks=nbins, 
      col="red",
      xlab='Daily # of steps', 
      ylim=c(0,12),
      main="Total Daily Number of Steps")
legend(500, 12, legend = c(paste("Mean =", round(meanSteps, 1)),
                          paste("Median =",round(mediSteps, 1))), bty = "n")
```

Our results show that by filling the NA with the median of the day interval changes the average slightly but does not affect the median significantly.

## Are there differences in activity patterns between weekdays and weekends?
```{r}
# CREATE FACTOR COLUMN FOR WEEKDAYS & WEEKENDS:
dayofweek   <- is.weekend(allvalues$date) 
dayofweek.f <- factor(dayofweek, labels=c("Weekday","Weekend"))
finaltable <- cbind(allvalues, dayofweek.f)

finaltable$interval <- strptime(sprintf("%04d", as.numeric(finaltable$interval)), format="%H%M")
finaltable$interval <- as.POSIXct(finaltable$interval, format="%H:%M:%S")

## GROUP BY INTERVALS & WEEKDAY/WEEKEND:
inter <- group_by(finaltable, interval, dayofweek.f)
aveSteps <- summarise(inter, meanSteps=mean(steps))

# PLOT RESULTS:
g <- ggplot(aveSteps, aes(interval, meanSteps)) + geom_point() + geom_line()
g <- g + labs(title = "Average # steps per day interval") 
g <- g + labs(x = "Day Interval", y = "Average # steps")
g <- g + facet_wrap(~dayofweek.f, ncol=1)#facet_grid(dayofweek.f ~ .) # + ylim(0,120)
g <- g + scale_x_datetime(date_breaks="2 hour", labels=date_format_tz("%H:%M", tz="EST"))

print(g)
```

Breaking the data down into week days and weekends show little difference other than the fact that the number of steps early in the morning (06:00am to 08:00am) is much lower during the weekends than on week days. This suggests that the person that wore the step counting device likes to sleep in on the weekends.












